Supplementary Materials for \"COATi: statistical pairwise alignment of protein coding sequences\

### 2024-06-07 alignpair_letter_source.tex
Draft of final version after acceptence

% now start line numbers
\linenumbers

@@

Bioinformatic pipelines need to be robust to variation in quality across genomic datasets because uncorrected errors in the alignment stage can lead to erroneous results in comparative and functional genomic studies \citep{estimates_schneider_2009, effect_fletcher_2010, hubisz2011error}.
While genomes for model organisms often get refined over many iterations and contain meticulously curated protein-coding sequences, 
genomes for non-model organisms might only receive partial curation and typically have lower quality sequences and annotations.
These genomes often lack the amount of sequencing data needed to fix artifacts, including missing exons, erroneous mutations, and indels \citep{jackman2018tigmint}.
%
When comparative and functional genomics studies include data from non-model organisms, care must be taken to identify and manage such artifacts; however,
current alignment methods are ill-equipped to handle common artifacts in genomic data, requiring costly curation practices that discard significant amounts of information.

To address current limitations of alignment software to accurately align protein-coding sequences, we present COATi, short for COdon-aware Alignment Transducer, a pairwise statistical aligner that incorporates evolutionary models for protein-coding sequences and is robust to artifacts present in modern genomic data sets. We find that COATi generates biologically reasonable alignments and downstream inferences when applied to empirical data. We also find that COATi generates more accurate alignments when applied to semi-empirical benchmarks.

@@

In order to compare COATi against other aligners on realistic datasets with known alignments, we developed a procedure to introduce realistic gap patterns into human-gorilla orthologous gene pairs that did not previously contain indels. We separated our 14,127 empirical sequence pairs into two sets: %
(1) 8261 sequence pairs that did not contain any gaps when aligned by COATi FST, Clustal\textOmega{}, MACSE, MAFFT, and PRANK, and (2) 5821 sequence pairs for which at least one of these five aligners added gaps.
(45 pairs that could not be aligned by at least one aligner were dropped.) We identified gap patterns from the sequence pairs that were aligned with gaps and randomly introduced them into the ungapped sequence pairs. We used an equal number of randomly sampled gap patterns from each aligner to produce a benchmark set of known alignments. The phases of gaps were preserved as were the spacings of any clusters of gaps. Segments of matches that were 96 nucleotides or longer were allowed to change length to accommodate the length of the ungapped sequence pair. (This criteria was recursively lowered if the gap pattern did not fit the ungapped sequence pair.)

For example, consider the gapped alignment represented by the CIGAR string
%
``170M 3D 10M 6I 102M'' % Can use \allowbreak{} here instead of space
%
applied to an ungapped alignment of human and gorilla sequences that are 300 nucleotides long. First, we modified the CIGAR string to insert flexible lengths for any match segment that is 96 or more nucleotides long while preserving phase. This results in a new CIGAR string of
%
``98M *M 3D 10M 6I 96M *M'',
%
where * represents the locations that have flexible length. Considering only matches and deletions, this CIGAR string has 207 fixed nucleotides, leaving 93 nucleotides in the human sequence to be allocated to the flexible locations. Since there are two locations of flexible length, we drew one random break point uniformly while maintaining phase, producing the final CIGAR string of
%
``98M 12M 3D 10M 6I 96M 81M'',
%
which was used to add gaps into the target, ungapped sequence pair.
To apply deletions, we removed the corresponding nucleotides from the gorilla sequence, and to apply insertions, we added nucleotides to the gorilla sequence by sampling codons from their stationary frequency while respecting gap phases. The human sequence was left unchanged.
%
We generated a benchmark of 8261 sequence pairs with known alignments using this strategy.

@@

COATi produced better alignments than Clustal\textOmega{}, MACSE, MAFFT, and PRANK when evaluated by our benchmark data set. It produced more best alignments, more perfect alignments, and less imperfect alignments (Tab.~\ref{table:comp}).
%
COATi was significantly more accurate (lower $d_{seq}$) at inferring the benchmark alignments compared to the other methods (Tab.~\ref{table:comp}). All p-values were significant according to the one-tailed, paired Wilcoxon signed-rank tests. (P-value of COATi vs PRANK was $5.1 \times 10^{-8}$, and p-values of COATi vs the other aligners were $< 2.2 \times 10^{-16}$). Notably, the average alignment error of the second best protocol was four times larger than COATi's. We also calculated the average distance of alignments between aligners and used principal coordinate analysis (PCoA) to project the resulting distance matrix into two dimensions (Fig.~\ref{fig:dseq-benchmarks}). COATi was the closest aligner to the benchmark, and other aligners tended to diverge from the benchmark in different directions. Because benchmark alignments were derived from gap patterns estimated by different aligners, we also performed separate PCoA analyses for every source of gap patterns (Supplementary Materials, Fig.~S6). While all aligners, except Clustal\textOmega{}, do exceptionally well on their own patterns, COATi does exceptionally well on every set of patterns.

@@

We have developed a statistical alignment program using finite-state transducers called COATi. On empirical data, COATi produced alignments that provided reasonable biological inferences. On semi-empirical benchmarks, COATi was more accurate than the other aligners/alignment strategies that we tested: Clustal\textOmega{}'s amino-acid model, MACSE's amino-acid aware nucleotide model, MAFFT's DNA model, and PRANK's codon model. Unlike standard codon or amino-acid alignment strategies, COATi supports aligning protein-coding sequences using all three phases of gaps. Consistent with \cite{taylor2004occurrence} and \cite{zhu2022profiling}, our COATi results indicate that only 41\% indels in protein-coding sequences are phase-3 indels between humans and gorillas. Therefore, alignment strategies that only support phase-3 gaps in protein-coding sequences are suboptimally aligning the 59\% of indels that are phase-1 and phase-2.

@@

We can look at distances between alignments generated by different aligners to understand the diversity of approaches used by our alignment protocols. Aside from the cluster of different COATi methods, we detected no other clusters in our analysis (Supplemental Materials Fig.~S7). Different aligners tended to spread out from one another and in different directions from the benchmark. This indicates that each protocol that we utilized approached alignment inference from a different section of ``aligner space''. While each aligner, except Clustal\textOmega{}, performed well on its own gap patterns, MACSE also did well on Clustal\textOmega{} and MAFFT gap patterns, and PRANK did well on Clustal\textOmega{} patterns. COATi did well on all patterns.

@@

COATi is under active development, and future work will enhance the multiple-sequence aligner to refine the initial alignment according to a scoring function consistent with COATi's FST model. We also plan on extending COATi to support more complex gap models, e.g.\ mixtures of single-nucleotide and triple-nucleotide indel models and weighting gap openings to reflect known selection on indel phases \citep{zhu2022profiling}. We also plan on improving its alignment sampling capabilities, as well as implement new models for aligning long-read sequences of genes against reference genomes. Our goal is to develop COATi into a user-friendly suite of tools that will allow researchers to analyze more data with higher accuracy and facilitate the study of important biological processes that shape genomic data.

@@

The source code for COATi, along with documentation, is freely available on GitHub: \url{https://github.com/CartwrightLab/coati} and is implemented in C++. COATi is released under an MIT license. Additional information, code, and workflows to replicate the analysis can be found on GitHub: \url{https://github.com/jgarciamesa/coati-testing} and \url{https://github.com/jgarciamesa/alignpair_letter}.

@@

This research was funded by NSF award DBI-1929850.
%
The authors would like to thank Profs.\ Marco Mangone, Ted Pavlic, Banu Oskan, Jay Taylor, and Jeremy Wideman for their helpful support on two separate PhD dissertation projects. The authors would also like to thank the associate editor and two reviewers for their helpful comments on earlier versions of this manuscript.

@@

\begin{figure}[ht]
    \centering%
    %\includegraphics{figures/fig-dseq-benchmarks.pdf}
    \par
    \caption{COATi's alignments were closer to the semi-empirical benchmark than other methods. A principal coordinates analysis (PCoA) of the average alignment distance ($d_{seq}$) between alignments generated by different methods.\label{fig:dseq-benchmarks}}
\end{figure}

@@

% Software comparison table
\begin{table}[ht]
\definecolor{bestcolor}{RGB}{230,230,230}%
\centering
\caption{COATi generates better alignments than other alignment algorithms on semi-empirical benchmarks.\textsuperscript{1}\label{table:comp}}
%
% Copied from tables/benchmarks.tex
% footnotes and \addlinespace are manually added
\begin{tabular}{rccccc}
  \toprule
 & COATi & Clustal\textOmega{} & MACSE & MAFFT & PRANK \\ 
  \midrule
Average Alignment Error ($d_{seq}$) & \cellcolor{bestcolor}0.23\% & 2.65\% & 1.37\% & 1.67\% & 1.02\% \\
\addlinespace 
  Number of Best\textsuperscript{2} Alignments & \cellcolor{bestcolor}7050 & 3988 & 5226 & 5918 & 6128 \\ 
  Number of Perfect\textsuperscript{2} Alignments & \cellcolor{bestcolor}6795 & 3954 & 5138 & 5840 & 5860 \\ 
  Number of Imperfect\textsuperscript{2} Alignments & \cellcolor{bestcolor} 850 & 3691 & 2507 & 1805 & 1785 \\
  \addlinespace 
  RMSE for K2P Distances & \cellcolor{bestcolor}0.000694 & 0.0920 & 0.0528 & 0.0561 & 0.0305 \\ 
  Overestimated K2P Distances & \cellcolor{bestcolor}10.3\% & 52.7\% & 38.0\% & 26.2\% & 25.6\% \\ 
  \addlinespace
  F\textsubscript{1} Score for Negative Selection & \cellcolor{bestcolor}99.8\% & 97.6\% & 98.4\% & 98.3\% & 99.1\% \\ 
  F\textsubscript{1} Score for Positive Selection & \cellcolor{bestcolor}97.8\% & 77.2\% & 84.9\% & 85.5\% & 91.8\% \\ 
   \bottomrule
   \addlinespace
\multicolumn{6}{p{40em}}{\textsuperscript{1}Total number of sequence pairs in the benchmark was 8,261.}\\
\multicolumn{6}{p{40em}}{\textsuperscript{2}Best alignments have the lowest $d_{seq}$ (including equivalent alignments). Perfect alignments have $d_{seq} = 0$ (including equivalent alignments). Imperfect alignments are alignments that are not perfect when at least one method found a perfect alignment. }\\
\end{tabular}
\end{table}

\clearpage
\newpage
\bibliographystyle{mbe}%
\setlength{\bibhang}{0pt}
%\bibliography{alignpair_letter.bib}
\begin{thebibliography}{43}
\providecommand{\natexlab}[1]{#1}

\bibitem[{Abascal et~al.(2010)Abascal, Zardoya, and
  Telford}]{abascal2010translatorx}
Abascal F, Zardoya R, and Telford MJ. 2010.
\newblock {TranslatorX}: multiple alignment of nucleotide sequences guided by
  amino acid translations.
\newblock \emph{Nucleic Acids Research} 38:W7--W13.

\bibitem[{Allauzen et~al.(2007)Allauzen, Riley, Schalkwyk, Skut, and
  Mohri}]{allauzen2007openfst}
Allauzen C, Riley M, Schalkwyk J, Skut W, and Mohri M. 2007.
\newblock Openfst: A general and efficient weighted finite-state transducer
  library.
\newblock In \emph{International Conference on Implementation and Application
  of Automata}, pages 11--23. Springer.

\bibitem[{Arvestad(1997)}]{arvestad1997aligning}
Arvestad L. 1997.
\newblock Aligning coding dna in the presence of frame-shift errors.
\newblock In A~Apostolico and J~Hein, editors, \emph{Combinatorial Pattern
  Matching}, pages 180--190. Berlin, Heidelberg: Springer Berlin Heidelberg.

\bibitem[{Bininda-Emonds(2005)}]{bininda2005transalign}
Bininda-Emonds O. 2005.
\newblock {transAlign}: using amino acids to facilitate the multiple alignment
  of protein-coding {DNA} sequences.
\newblock \emph{BMC Bioinformatics} 6:1--6.

\bibitem[{Blackburne and Whelan(2011)}]{metrics_blackburne_whelan_2011}
Blackburne BP and Whelan S. 2011.
\newblock {Measuring the distance between multiple sequence alignments}.
\newblock \emph{Bioinformatics} 28:495--502.

\bibitem[{Bradley and Holmes(2007)}]{bradley2007transducers}
Bradley RK and Holmes I. 2007.
\newblock Transducers: an emerging probabilistic framework for modeling indels
  on trees.
\newblock \emph{Bioinformatics} 23.

@@

\bibitem[{Hein(1994)}]{hein1994algorithm}
Hein J. 1994.
\newblock An algorithm combining dna and protein alignment.
\newblock \emph{Journal of Theoretical Biology} 167:169--174.

@@

\bibitem[{Holmes and Bruno(2001)}]{holmes2001evolutionary}
Holmes I and Bruno WJ. 2001.
\newblock Evolutionary {HMMs}: a bayesian approach to multiple alignment.
\newblock \emph{Bioinformatics} 17:803--820.

@@

\bibitem[{Jackman et~al.(2018)Jackman, Coombe, Chu, Warren, Vandervalk, Yeo,
  Xue, Mohamadi, Bohlmann, Jones et~al.}]{jackman2018tigmint}
Jackman SD, Coombe L, Chu J, Warren RL, Vandervalk BP, Yeo S, Xue Z, Mohamadi
  H, Bohlmann J, Jones SJ, et~al. 2018.
\newblock Tigmint: correcting assembly errors using linked reads from large
  molecules.
\newblock \emph{BMC Bioinformatics} 19:1--10.

\bibitem[{Katoh and Standley(2013)}]{katoh2013mafft}
Katoh K and Standley DM. 2013.
\newblock {MAFFT} multiple sequence alignment software version 7: improvements
  in performance and usability.
\newblock \emph{Molecular Biology and Evolution} 30:772--780.

@@

\bibitem[{Mohri et~al.(2005)Mohri, Pereira, and Riley}]{mohri2005weighted}
Mohri M, Pereira F, and Riley M. 2005.
\newblock Weighted automata in text and speech processing.

@@

\bibitem[{Pamilo and Bianchi(1993)}]{Pamilo1993}
Pamilo P and Bianchi NO. 1993.
\newblock {Evolution of the Zfx and Zfy genes: rates and interdependence
  between the genes.}
\newblock \emph{Molecular Biology and Evolution} 10:271--281.

\bibitem[{Paradis and Schliep(2019)}]{paradis2019ape}
Paradis E and Schliep K. 2019.
\newblock ape 5.0: an environment for modern phylogenetics and evolutionary
  analyses in {R}.
\newblock \emph{Bioinformatics} 35:526--528.

@@

\bibitem[{Ranwez et~al.(2018)Ranwez, Douzery, Cambon, Chantret, and
  Delsuc}]{ranwez_macse_2018}
Ranwez V, Douzery EJP, Cambon C, Chantret N, and Delsuc F. 2018.
\newblock {{MACSE} v2: toolkit for the alignment of coding sequences accounting
  for frameshifts and stop codons}.
\newblock \emph{Molecular Biology and Evolution} 35:2582--2584.

@@

\bibitem[{Saitou and Nei(1987)}]{saitou1987neighbor}
Saitou N and Nei M. 1987.
\newblock The neighbor-joining method: a new method for reconstructing
  phylogenetic trees.
\newblock \emph{Molecular Biology and Evolution} 4:406--425.

@@

\bibitem[{Silvestre-Ryan et~al.(2021)Silvestre-Ryan, Wang, Sharma, Lin, Shen,
  Dider, and Holmes}]{silvestre2021machine}
Silvestre-Ryan J, Wang Y, Sharma M, Lin S, Shen Y, Dider S, and Holmes I. 2021.
\newblock {Machine Boss}: rapid prototyping of bioinformatic automata.
\newblock \emph{Bioinformatics} 37:29--35.

>>>>

% now start line numbers
%\linenumbers

@@

Bioinformatic pipelines need to be robust to variation in quality across genomic datasets because uncorrected errors in the alignment stage can lead to erroneous results in comparative and functional genomic studies \citep{estimates_schneider_2009, effect_fletcher_2010, hubisz2011error}.
While genomes for model organisms are often refined over many iterations and contain meticulously curated protein-coding sequences, 
genomes for non-model organisms might only receive partial curation and typically have lower quality sequences and annotations.
These genomes often lack the amount of sequencing data needed to fix artifacts, including missing exons, erroneous mutations, and indels \citep{jackman2018tigmint}.
%
When comparative and functional genomics studies include data from non-model organisms, care must be taken to identify and manage such artifacts; however,
current alignment methods are ill-equipped to handle common artifacts in genomic data, requiring costly curation practices that discard significant amounts of information.

To address current limitations of alignment software to accurately align protein-coding sequences, we present COATi, short for COdon-aware Alignment Transducer, a pairwise statistical aligner that incorporates evolutionary models for protein-coding sequences and is robust to artifacts present in modern genomic datasets. We find that COATi generates biologically reasonable alignments and downstream inferences when applied to empirical data. We also find that COATi generates more accurate alignments when applied to a semi-empirical benchmark dataset.

@@

In order to compare COATi against other aligners on realistic datasets with known alignments, we developed a procedure to introduce realistic gap patterns into human-gorilla orthologous gene pairs that did not previously contain indels. We separated our 14,127 empirical sequence pairs into two sets: %
(1) 8261 sequence pairs that did not contain any gaps when aligned by COATi FST, Clustal\textOmega{}, MACSE, MAFFT, and PRANK, and (2) 5821 sequence pairs for which at least one of these five aligners added gaps.
(45 pairs that could not be aligned by at least one aligner were dropped.) We identified gap patterns from the sequence pairs that were aligned with gaps and randomly introduced them into the ungapped sequence pairs. We used an equal number of randomly sampled gap patterns from each aligner to produce a benchmark dataset of known alignments. The phases of gaps were preserved as were the spacings of any clusters of gaps. Segments of matches that were 96 nucleotides or longer were allowed to change length to accommodate the length of the ungapped sequence pair. (This criteria was recursively lowered if the gap pattern did not fit the ungapped sequence pair.)

For example, consider the gapped alignment represented by the CIGAR string
%
``170M 3D 10M 6I 102M'' % Can use \allowbreak{} here instead of space
%
applied to an ungapped alignment of human and gorilla sequences that are 300 nucleotides long. First, we modified the CIGAR string to insert flexible lengths for any match segment that is 96 or more nucleotides long while preserving phase. This results in a new CIGAR string of
%
``98M *M 3D 10M 6I 96M *M'',
%
where * represents the locations that have flexible length. Considering only matches and deletions, this CIGAR string has 207 fixed nucleotides, leaving 93 nucleotides in the human sequence to be allocated to the flexible locations. Since there are two locations of flexible length, we drew one random break point uniformly while maintaining phase, producing the final CIGAR string of
%
``98M 12M 3D 10M 6I 96M 81M'',
%
which was used to add gaps into the target, ungapped sequence pair.
To apply deletions, we removed the corresponding nucleotides from the gorilla sequence, and to apply insertions, we added nucleotides to the gorilla sequence by sampling codons from their stationary frequency while respecting gap phases. The human sequence was left unchanged.
%
We generated a benchmark dataset of 8261 sequence pairs with known alignments using this strategy.

@@

COATi produced better alignments than Clustal\textOmega{}, MACSE, MAFFT, and PRANK when evaluated by our benchmark dataset. It produced more best alignments, more perfect alignments, and less imperfect alignments (Tab.~\ref{table:comp}).
%
COATi was significantly more accurate (lower $d_{seq}$) at inferring the benchmark alignments compared to the other methods (Tab.~\ref{table:comp}). All p-values were significant according to the one-tailed, paired Wilcoxon signed-rank tests. (P-value of COATi vs PRANK was $5.1 \times 10^{-8}$, and p-values of COATi vs the other aligners were $< 2.2 \times 10^{-16}$). Notably, the average alignment error of the second best protocol was four times larger than COATi's. We also calculated the average distance of alignments between aligners and used principal coordinate analysis (PCoA) to project the resulting distance matrix into two dimensions (Fig.~\ref{fig:dseq-benchmarks}). COATi was the closest aligner to the benchmark dataset, and other aligners tended to diverge from the benchmark dataset in different directions. Because benchmark alignments were derived from gap patterns estimated by different aligners, we also performed separate PCoA analyses for every source of gap patterns (Supplementary Materials, Fig.~S6). While all aligners, except Clustal\textOmega{}, do exceptionally well on their own patterns, COATi does exceptionally well on every set of patterns.

@@

We have developed a statistical alignment program using finite-state transducers called COATi. On empirical data, COATi produced alignments that provided reasonable biological inferences. On a semi-empirical benchmark dataset, COATi was more accurate than the other aligners/alignment strategies that we tested: Clustal\textOmega{}'s amino-acid model, MACSE's amino-acid aware nucleotide model, MAFFT's DNA model, and PRANK's codon model. Unlike standard codon or amino-acid alignment strategies, COATi supports aligning protein-coding sequences using all three phases of gaps. Consistent with \cite{taylor2004occurrence} and \cite{zhu2022profiling}, our COATi results indicate that only 41\% indels in protein-coding sequences are phase-3 indels between humans and gorillas. Therefore, alignment strategies that only support phase-3 gaps in protein-coding sequences are suboptimally aligning the 59\% of indels that are phase-1 and phase-2.

@@

We can look at distances between alignments generated by different aligners to understand the diversity of approaches used by our alignment protocols. Aside from the cluster of different COATi methods, we detected no other clusters in our analysis (Supplemental Materials Fig.~S7). Different aligners tended to spread out from one another and in different directions from the benchmark dataset. This indicates that each protocol that we utilized approached alignment inference from a different section of ``aligner space''. While each aligner, except Clustal\textOmega{}, performed well on its own gap patterns, MACSE also did well on Clustal\textOmega{} and MAFFT gap patterns, and PRANK did well on Clustal\textOmega{} patterns. COATi did well on all patterns.

@@

COATi is under active development, and future work will enhance the multiple-sequence aligner to refine the initial alignment according to a scoring function consistent with COATi's FST model. We also plan on extending COATi to support more complex gap models, e.g.\ mixtures of single-nucleotide and triple-nucleotide indel models and weighting gap openings to reflect known selection on indel phases \citep{zhu2022profiling}. We also plan on improving its alignment sampling capabilities, as well as implementing new models for aligning long-read sequences of genes against reference genomes. Our goal is to develop COATi into a user-friendly suite of tools that will allow researchers to analyze more data with higher accuracy and facilitate the study of important biological processes that shape genomic data.

@@

The source code for COATi, along with documentation, is freely available on GitHub: \nolinkurl{https://github.com/CartwrightLab/coati} (DOI: \nolinkurl{10.5281/zenodo.11499800}) and is implemented in C++. COATi is released under an MIT license. Additional information, code, and workflows to replicate the analysis can be found on GitHub: \nolinkurl{https://github.com/jgarciamesa/coati-testing} (DOI: \nolinkurl{10.5281/zenodo.11515933}) and \nolinkurl{https://github.com/jgarciamesa/alignpair_letter} (DOI: \nolinkurl{10.5281/zenodo.11512411}).

@@

This research was funded by NSF award DBI-1929850.
%
The authors would like to thank Profs.\ Marco Mangone, Ted Pavlic, Banu Ozkan, Jay Taylor, and Jeremy Wideman for their helpful support on two separate PhD dissertation projects. The authors would also like to thank the associate editor and two reviewers for their helpful comments on earlier versions of this manuscript.

@@

\begin{figure}[ht]
    \centering%
    %\includegraphics{figures/fig-dseq-benchmarks.pdf}
    \par
    \caption{COATi's alignments were closer to the semi-empirical benchmark dataset than other methods. A principal coordinates analysis (PCoA) of the average alignment distance ($d_{seq}$) between alignments generated by different methods.\label{fig:dseq-benchmarks}}
\end{figure}

@@

% Software comparison table
\begin{table}[ht]
\definecolor{bestcolor}{RGB}{230,230,230}%
\centering
\caption{COATi generates better alignments than other alignment algorithms on a semi-empirical benchmark dataset.\textsuperscript{1}\label{table:comp}}
%
% Copied from tables/benchmarks.tex
% footnotes and \addlinespace are manually added
\begin{tabular}{rccccc}
  \toprule
 & COATi & Clustal\textOmega{} & MACSE & MAFFT & PRANK \\ 
  \midrule
Average Alignment Error ($d_{seq}$) & \cellcolor{bestcolor}0.23\% & 2.65\% & 1.37\% & 1.67\% & 1.02\% \\
\addlinespace 
  Number of Best\textsuperscript{2} Alignments & \cellcolor{bestcolor}7050 & 3988 & 5226 & 5918 & 6128 \\ 
  Number of Perfect\textsuperscript{2} Alignments & \cellcolor{bestcolor}6795 & 3954 & 5138 & 5840 & 5860 \\ 
  Number of Imperfect\textsuperscript{2} Alignments & \cellcolor{bestcolor} 850 & 3691 & 2507 & 1805 & 1785 \\
  \addlinespace 
  RMSE for K2P Distances & \cellcolor{bestcolor}0.000694 & 0.0920 & 0.0528 & 0.0561 & 0.0305 \\ 
  Overestimated K2P Distances & \cellcolor{bestcolor}10.3\% & 52.7\% & 38.0\% & 26.2\% & 25.6\% \\ 
  \addlinespace
  F\textsubscript{1} Score for Negative Selection & \cellcolor{bestcolor}99.8\% & 97.6\% & 98.4\% & 98.3\% & 99.1\% \\ 
  F\textsubscript{1} Score for Positive Selection & \cellcolor{bestcolor}97.8\% & 77.2\% & 84.9\% & 85.5\% & 91.8\% \\ 
   \bottomrule
   \addlinespace
\multicolumn{6}{p{40em}}{\textsuperscript{1}Total number of sequence pairs in the benchmark dataset was 8,261.}\\
\multicolumn{6}{p{40em}}{\textsuperscript{2}Best alignments have the lowest $d_{seq}$ (including equivalent alignments). Perfect alignments have $d_{seq} = 0$ (including equivalent alignments). Imperfect alignments are alignments that are not perfect when at least one method found a perfect alignment. }\\
\end{tabular}
\end{table}

\clearpage
\newpage
\bibliographystyle{mbe}%
\setlength{\bibhang}{0pt}
%\bibliography{alignpair_letter.bib}
\begin{thebibliography}{44}
\providecommand{\natexlab}[1]{#1}

\bibitem[{Abascal et~al.(2010)Abascal, Zardoya, and
  Telford}]{abascal2010translatorx}
Abascal F, Zardoya R, and Telford MJ. 2010.
\newblock {TranslatorX}: Multiple alignment of nucleotide sequences guided by
  amino acid translations.
\newblock \emph{Nucleic Acids Research} 38:W7--W13.

\bibitem[{Allauzen et~al.(2007)Allauzen, Riley, Schalkwyk, Skut, and
  Mohri}]{allauzen2007openfst}
Allauzen C, Riley M, Schalkwyk J, Skut W, and Mohri M. 2007.
\newblock {OpenFst}: A general and efficient weighted finite-state transducer
  library.
\newblock In \emph{International Conference on Implementation and Application
  of Automata}, pages 11--23. Springer.

\bibitem[{Arvestad(1997)}]{arvestad1997aligning}
Arvestad L. 1997.
\newblock Aligning coding {DNA} in the presence of frame-shift errors.
\newblock In A~Apostolico and J~Hein, editors, \emph{Combinatorial Pattern
  Matching}, pages 180--190. Berlin, Heidelberg: Springer Berlin Heidelberg.

\bibitem[{Bininda-Emonds(2005)}]{bininda2005transalign}
Bininda-Emonds O. 2005.
\newblock {transAlign}: Using amino acids to facilitate the multiple alignment
  of protein-coding {DNA} sequences.
\newblock \emph{BMC Bioinformatics} 6:1--6.

\bibitem[{Blackburne and Whelan(2011)}]{metrics_blackburne_whelan_2011}
Blackburne BP and Whelan S. 2011.
\newblock Measuring the distance between multiple sequence alignments.
\newblock \emph{Bioinformatics} 28:495--502.

\bibitem[{Bradley and Holmes(2007)}]{bradley2007transducers}
Bradley RK and Holmes I. 2007.
\newblock Transducers: An emerging probabilistic framework for modeling indels
  on trees.
\newblock \emph{Bioinformatics} 23.

@@

\bibitem[{Hein(1994)}]{hein1994algorithm}
Hein J. 1994.
\newblock An algorithm combining {DNA} and protein alignment.
\newblock \emph{Journal of Theoretical Biology} 167:169--174.

@@

\bibitem[{Holmes(2020)}]{Holmes2020}
Holmes I. 2020.
\newblock A model of indel evolution by finite-state, continuous-time machines.
\newblock \emph{Genetics} 216:1187--1204.

\bibitem[{Holmes and Bruno(2001)}]{holmes2001evolutionary}
Holmes I and Bruno WJ. 2001.
\newblock Evolutionary {HMMs}: A {Bayesian} approach to multiple alignment.
\newblock \emph{Bioinformatics} 17:803--820.

@@

\bibitem[{Jackman et~al.(2018)Jackman, Coombe, Chu, Warren, Vandervalk, Yeo,
  Xue, Mohamadi, Bohlmann, Jones et~al.}]{jackman2018tigmint}
Jackman SD, Coombe L, Chu J, Warren RL, Vandervalk BP, Yeo S, Xue Z, Mohamadi
  H, Bohlmann J, Jones SJ, et~al. 2018.
\newblock Tigmint: Correcting assembly errors using linked reads from large
  molecules.
\newblock \emph{BMC Bioinformatics} 19:1--10.

\bibitem[{Katoh and Standley(2013)}]{katoh2013mafft}
Katoh K and Standley DM. 2013.
\newblock {MAFFT} multiple sequence alignment software version 7: Improvements
  in performance and usability.
\newblock \emph{Molecular Biology and Evolution} 30:772--780.

@@

\bibitem[{Mohri et~al.(2005)Mohri, Pereira, and Riley}]{mohri2005weighted}
Mohri M, Pereira F, and Riley M. 2005.
\newblock Weighted automata in text and speech processing.
\newblock arXiv https://arxiv.org/abs/cs/0503077.

@@

\bibitem[{Pamilo and Bianchi(1993)}]{Pamilo1993}
Pamilo P and Bianchi NO. 1993.
\newblock {Evolution of the {Zfx} and {Zfy} genes: rates and interdependence
  between the genes.}
\newblock \emph{Molecular Biology and Evolution} 10:271--281.

\bibitem[{Paradis and Schliep(2019)}]{paradis2019ape}
Paradis E and Schliep K. 2019.
\newblock {ape} 5.0: an environment for modern phylogenetics and evolutionary
  analyses in {R}.
\newblock \emph{Bioinformatics} 35:526--528.

@@

\bibitem[{Ranwez et~al.(2018)Ranwez, Douzery, Cambon, Chantret, and
  Delsuc}]{ranwez_macse_2018}
Ranwez V, Douzery EJP, Cambon C, Chantret N, and Delsuc F. 2018.
\newblock {{MACSE} v2: Toolkit for the alignment of coding sequences accounting
  for frameshifts and stop codons}.
\newblock \emph{Molecular Biology and Evolution} 35:2582--2584.

@@

\bibitem[{Saitou and Nei(1987)}]{saitou1987neighbor}
Saitou N and Nei M. 1987.
\newblock The neighbor-joining method: A new method for reconstructing
  phylogenetic trees.
\newblock \emph{Molecular Biology and Evolution} 4:406--425.

@@

\bibitem[{Silvestre-Ryan et~al.(2021)Silvestre-Ryan, Wang, Sharma, Lin, Shen,
  Dider, and Holmes}]{silvestre2021machine}
Silvestre-Ryan J, Wang Y, Sharma M, Lin S, Shen Y, Dider S, and Holmes I. 2021.
\newblock {Machine Boss}: Rapid prototyping of bioinformatic automata.
\newblock \emph{Bioinformatics} 37:29--35.
